Polynomial phase estimation by phase unwrapping

### 2013-02-19 paper.tex
added now new data.  Added to paper

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  Under mild conditions on the distribution of the noise we describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed.  No `small noise' assumption (common in the literature) is required in our analysis.  As a result the LSU estimator will be accurate, even for very noisy signals, provided that the number of observations is sufficiently large.  The LSU estimator can be computed in a reasonable amount of time for data sets of moderate size using existing general purpose algorithms from algebraic number theory.  Monte-carlo simualtions assert the excellent statistical performance of the LSU estimator when compared to existing state-of-the-art estimators.  A key feature is that the LSU estimator functions accurately over a far wider range of parameters than most existing estimators.
\end{abstract}

@@

Polynomial phase signals arise in fields including radar, sonar, geophysics, biology, and radio communication~\cite{Hlawatsch_lin_quad_time_freq_spmag_1992,Ridleyspeechpolyphase1989, Suga_1975_bats_echolocation, Moss_2005echolocation,Angeby_estimating_2000}. %In radar and sonar applications polynomial phase signals arise when acquiring radial velocity and acceleration (and higher order motion descriptors) of a target from a reflected signal, and also in continuous wave radar and low probability of intercept radar~\cite{Levanon_Radar_signals_2004}.  In biology, polynomial phase signals are used to describe the sounds emitted by bats and dolphins for echo location \citep{Suga_1975_bats_echolocation, Moss_2005echolocation}.  
A polynomial phase signal of order $m$ is a function of the form
\[
s(t) = e^{2\pi j y(t)},
\]
where $j = \sqrt{-1}$, and $t$ is a real number and 
\[
y(t) = \tilde{\mu}_0 +\tilde{\mu}_1 t + \tilde{\mu}_2 t^2 + \dots \tilde{\mu}_m t^m
\]
is a polynomial of order $m$.  In practice the signal is typically sampled at discrete points in `time', $t$. In this paper we only consider uniform sampling, where the gap between consecutive samples is constant. In this case we can always consider the samples to be taken at some set of consecutive integers and our sampled polynomial phase signal takes the form
\[
s_n = s(n) = e^{2\pi j y(n)},
\] 
where $n$ is an integer.  Of practical importance is the estimation of the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from a number, say $N$, of observations of the noisy sampled signal
\begin{equation}\label{eq:Y_nsamplednoisey}
Y_n = \rho s_n + X_n,
\end{equation}
where $\rho$ is a positive real number representing the (usually unknown) signal amplitude and $\{X_n, n \in \ints\}$ is a sequence of complex noise variables. In order to ensure identifiability it is necessary to restrict the $m+1$ coefficients to a region of $m+1$ dimensional Euclidean space $\reals^{m+1}$ called an \emph{identifiable region}.  It was shown in \cite{McKilliam2009IndentifiabliltyAliasingPolyphase} that an identifiable region tessellates a particular $m+1$ dimensional lattice.  We discuss this further in Section \ref{sec:ident_aliasing}.

An obvious estimator of the unknown coefficients is the least squares estimator (LSE).  This is also the maximum likelihood estimator (MLE) when the noise sequence $\{X_n\}$ is white and Gaussian.  When $m=0$ (phase estimation) or $m=1$ (frequency estimation) the LSE is an effective approach, being both computationally efficient and statistically accurate \cite{Quinn2009_dasp_phase_only_information_loss,Hannan1973,Quinn2001,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam2010thesis}. When $m \geq 2$ the computational complexity of the LSE is large~\cite{McKilliam2010thesis,Abatzoglou_ml_chirp_1986}. For this reason many authors have considered alternative approaches to polynomial phase estimation. These can loosely be grouped into two classes: estimators based on `multilinear transforms', such as the high order ambiguity function (HAF)~\cite{Peleg_DPT_1995,Peleg1991_est_class_PPS_1991,Porat_asympt_HAF_DPT_1996,Farquharson_another_poly_est_2005,Boashash_Oshea_1994}, and cubic phase function (CPF)~\cite{Oshea_cpf_2004,Oshea_cpf_2002_letter,Djurovic_haf_cpf_2012}; and estimators based on phase unwrapping, such as Kitchen's unwrapping estimator~\cite{Kitchen_polyphase_unwrapping_1994}, the estimator of Djuric and Kay~\cite{Djuric_phase_unwrap_chirp_1990}, and Morelande's Bayesian unwrapping estimator~\cite{Morelande_bayes_unwrapping_2009_tsp}.

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  
A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999} that exactly computes a nearest lattice point, and \emph{Babai's nearest plane algorithm}~\cite{Babai1986} that only produces an approximation. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for small $N$ less than about $50$.  Babai's algorithm is computationally simple, but its statistical performance was observed to be comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and may admit fast nearest point algorithms.  This has been studied in~\cite{McKilliam2010thesis} where polynomial time algorithms were found that compute the nearest point exactly.  Although polynomial time, the algorithms are still computationally demanding in practice.  Nevertheless fast (exact or approximate) algorithms may exist for these special lattices, that are yet to be discovered.  In this paper we apply another general purpose approximate nearest point algorithm, called the $K$-best method~\cite{Zhan2006_K_best_sphere_decoder} for approximating the LSU estimator.  We show that this provides near sphere decoder (exact) performance, but can be computed in a reasonable amount of time if $N$ is not too large (approximately $N$ less than $1000$).  %The $K$-best method is motivated by sequential decoding algorithms and, in the context of polynomial phase estimation, bares some resemblance to the Bayesian unwrapping algorithm studied by Morelande~\cite{Morelande_bayes_unwrapping_2009_tsp}.
%We show by Montecarlo simulation that the LSU estimator approximated using the $K$-best method is statistically superior to state-of-the-art estimators from the literature including the HAF and CPF.  We also compare the computational performance of the estimators and find that BLERG.

@@

The analysis of most existing polynomial phase estimators make what is known as a `small noise' assumption~\cite{Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990}.  This amounts to assuming that that estimator $\hat{\mu}_k$ of the $k$th coefficient is such that $N^{k}(\tilde{\mu}_k -  \hat{\mu}_k) \rightarrow 0$ in probability as $N \rightarrow \infty$.  This assumption is used to justify the use of first and second order expansions~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats}, typically called a `perturbation analysis' in this literature~\cite{Peleg_linear_FM,Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004}, to obtain expressions for the asymptotic mean square error of the estimator.  The `small noise' assumption is usually justified by Monte-Carlo simulations showing the asymptotic mean square error obtained by the perturbation analysis to be accurate provided that the signal to noise ratio (SNR) is larger than a `threshold' value (see the simulations in Section~\ref{sec:simulations} for examples).  Typically one observes the threshold to occur at smaller values of SNR for larger $N$.  A problem with the `small noise' assumption is that it is not know that the threshold continues to occur at increasingly smaller values of SNR as $N$ increases or if it infact converges to some finite (but perhaps small) value of SNR.  The consequence of the later being that the estimator will be innaccurate if the SNR is smaller than this finite value, regardless of the number of observations $N$ obtained.

An analysis of the HAF estimator under the assumption that the noise $\{X_n\}$ is white and Gaussian and without the `small noise' assumption has been made in~\cite{Porat_asympt_HAF_DPT_1996} (see inparticular the proof of Lemma~1 and the remark that follows).  To date, the literature does not contain a corresponding analysis of any estimator based on phase unwrapping.  This has apparently lead to the misconception that phase unwrapping estimators are only accurate with the noise is `small' in some sense.  This misconception would appear to be motivated by early papers such as~\cite{Tretter1985,Kay1989} where it was observed that the noise occuring in the phase of the signal only looks `Gaussian' when the complex noise producing it has small variance.  The results in this paper make it clear that phase unwrapping estimators can be accurate even at very low signal to noise ratios.  The information contained in the amplitude is not essential for accurate estimation of polynomial phase signals, and can be discarded with only small loss.  For example, in the case that $\{X_n\}$ is white and Gaussian, our results show that the asymptotic variance of the LSU estimator converges to the Cramer-Rao lower bound (CRB) as the SNR becomes large, and converges to approximately 3 times the CRB as the SNR approaches zero.  Related results have been presented previously~\cite{Quinn2009_dasp_phase_only_information_loss}.  It is worth noting that this factor of 3 times the CRB at low SNR is better than results obtained with estimators based on multi-linear transforms, such as the HAF and CPF, for which the ratio between the asymptotic variance and the CRB diverges as the SNR approaches zero~\cite{Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004,Djurovic_haf_cpf_2012}.

%Along with strong consistency, we also assert, under similarly mild conditions, that the LSU estimator is asymptotically normal. %In the case where the noise is normally distributed the asymptotic variance of the LSU estimator is close to the Cramer-Rao bound at high signal to noise ratio (SNR) and is close to $\tfrac{2}{3}\pi$ times that Cramer-Rao bound at low (SNR).  
%Space restrictions make it impossible for us to include a proof of asymptotic normality in this paper.

BLERG: Appears to be the first estimator for which theoretical results have been derived and which works over the entire identifiable region of parameters.

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was also derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to properly understand the statistical properties of polynomial phase estimators.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise $\{X_n\}$. The proof of strong consistency is given in Section~\ref{sec:strongconstproof}.  Section \ref{sec:simulations} describes the results of Monte Carlo simulations that compare the performance of the LSU estimator with some existing estimators.  These simulations agree with the derived asymptotic properties.

@@

@@

It may be helpful to observe Figures~\ref{circstatplot_line} to~\ref{fig:circstatplot_cube}.  In these, the phase (divided by $2\pi$) of two distinct polynomial phase signals is plotted on the left, and on the right the principal component of the phase (also divided by $2\pi$) is plotted. The circles display the samples at the integers. Note that the samples of the principal components intersect.  The corresponding polynomial phase signals are aliases.

@@

The comparitively small set of coefficients suitable for the HAF and CPF estimators has been studied previously~\cite{Angeby_PPS_aliasing_2000,Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a significant limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Zhou and Wang~\cite{Zhou_lag_diverity_PPT_1997} describe a method based on the extended Euclidean algorithm that enables the HAF estimator to operate on the entire identifiable region $B$.  Unfortunately this comes at the price of statistical accuracy.  Figure~\ref{plot:m3bigzw} shows the performance of the Zhou and Wang estimator when the coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$, the same as in Figure~\ref{plot:m3big}.  The threshold for the estimator is approximately~\unit[29]{dB}.  There does not appear to exist (atleast in the literature) a similar method for increasing the range of coefficients for the CPF estimator.

@@

\begin{figure}[p] 
   	\centering 
  		\includegraphics{code/data/plot567-1.mps} 
  		\caption{BLERG} 
  		\label{plot:polyest} 
 \end{figure}

\begin{figure}[p] 
   	\centering 
  		\includegraphics{code/data/plot567-2.mps} 
  		\caption{BLERG} 
  		\label{plot:polyest} 
 \end{figure}

\begin{figure}[p] 
   	\centering 
  		\includegraphics{code/data/plot567-3.mps} 
  		\caption{BLERG} 
  		\label{plot:polyest} 
 \end{figure}

>>>>

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  The LSU estimator can be computed in a reasonable amount of time for data sets of moderate size using existing general purpose algorithms from algebraic number theory.  Under mild conditions on the distribution of the noise we describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed.  No `small noise' assumption (common in the literature) is required in our analysis.  As a result the LSU estimator will be accurate, even for very noisy signals, provided that the number of observations is sufficiently large.  A key feature is that the LSU estimator functions accurately over a far wider range of parameters than many popular existing estimators.  Monte-carlo simualtions assert the excellent statistical performance of the LSU estimator when compared to existing state-of-the-art estimators and support our theoretical results.
\end{abstract}

@@

Polynomial phase signals arise in fields including radar, sonar, geophysics, biology, and radio communication~\cite{Hlawatsch_lin_quad_time_freq_spmag_1992,Ridleyspeechpolyphase1989, Suga_1975_bats_echolocation, Moss_2005echolocation,Angeby_estimating_2000}. %In radar and sonar applications polynomial phase signals arise when acquiring radial velocity and acceleration (and higher order motion descriptors) of a target from a reflected signal, and also in continuous wave radar and low probability of intercept radar~\cite{Levanon_Radar_signals_2004}.  In biology, polynomial phase signals are used to describe the sounds emitted by bats and dolphins for echo location \citep{Suga_1975_bats_echolocation, Moss_2005echolocation}.  
A polynomial phase signal of order $m$ is a function of the form
\[
s(t) = e^{2\pi j y(t)},
\]
where $j = \sqrt{-1}$, and $t$ is a real number and 
\[
y(t) = \tilde{\mu}_0 +\tilde{\mu}_1 t + \tilde{\mu}_2 t^2 + \dots \tilde{\mu}_m t^m
\]
is a polynomial of order $m$.  In practice the signal is typically sampled at discrete points in `time', $t$. In this paper we only consider uniform sampling, where the gap between consecutive samples is constant. In this case we can always consider the samples to be taken at some set of consecutive integers and our sampled polynomial phase signal takes the form
\[
s_n = s(n) = e^{2\pi j y(n)},
\] 
where $n$ is an integer.  Of practical importance is the estimation of the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from a number, say $N$, of observations of the noisy sampled signal
\begin{equation}\label{eq:Y_nsamplednoisey}
Y_n = \rho s_n + X_n \qquad n = 1, \dots, N,
\end{equation}
where $\rho$ is a positive real number representing the (usually unknown) signal amplitude and $\{X_n, n \in \ints\}$ is a sequence of complex noise variables. In order to ensure identifiability it is necessary to restrict the $m+1$ coefficients to a region of $m+1$ dimensional Euclidean space $\reals^{m+1}$ called an \emph{identifiable region}.  It was shown in \cite{McKilliam2009IndentifiabliltyAliasingPolyphase} that an identifiable region tessellates a particular $m+1$ dimensional lattice.  We discuss this further in Section \ref{sec:ident_aliasing}.

An obvious estimator of the unknown coefficients is the least squares estimator (LSE).  This is also the maximum likelihood estimator (MLE) when the noise sequence $\{X_n\}$ is white and Gaussian.  When $m=0$ (phase estimation) or $m=1$ (frequency estimation) the LSE is an effective approach, being both computationally efficient and statistically accurate \cite{Quinn2009_dasp_phase_only_information_loss,Hannan1973,Quinn2001,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam2010thesis}. When $m \geq 2$ the computational complexity of the LSE is large~\cite{Oshea_cpf_2004,McKilliam2010thesis,Abatzoglou_ml_chirp_1986}. For this reason many authors have considered alternative approaches to polynomial phase estimation. These can loosely be grouped into two classes: estimators based on `multilinear transforms', such as the high order ambiguity function (HAF)~\cite{Peleg_DPT_1995,Peleg1991_est_class_PPS_1991,Porat_asympt_HAF_DPT_1996,Farquharson_another_poly_est_2005,Boashash_Oshea_1994}, and cubic phase function (CPF)~\cite{Oshea_cpf_2004,Oshea_cpf_2002_letter,Djurovic_haf_cpf_2012}; and estimators based on phase unwrapping, such as Kitchen's unwrapping estimator~\cite{Kitchen_polyphase_unwrapping_1994}, the estimator of Djuric and Kay~\cite{Djuric_phase_unwrap_chirp_1990}, and Morelande's Bayesian unwrapping estimator~\cite{Morelande_bayes_unwrapping_2009_tsp}.

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator~\cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam2010thesis}.  It was shown in \cite{McKilliam2009asilomar_polyest_lattice, McKilliamFrequencyEstimationByPhaseUnwrapping2009} that the LSU estimator can be represented as a \emph{nearest lattice point problem}~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  
A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder}~\cite{Pohst_sphere_decoder_1981,Viterbo_sphere_decoder_1999} that exactly computes a nearest lattice point, and \emph{Babai's nearest plane algorithm}~\cite{Babai1986} that only produces an approximation. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for small $N$ less than about $60$.  Babai's algorithm is computationally simple, but its statistical performance was observed to be comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and may admit fast nearest point algorithms.  This has been studied in~\cite{McKilliam2010thesis} where polynomial time algorithms were found that compute the nearest point exactly.  Although polynomial time, the algorithms are still computationally demanding in practice.  Nevertheless fast (exact or approximate) algorithms may exist for these special lattices, that are yet to be discovered.  In this paper we apply another general purpose approximate nearest point algorithm, called the $K$-best method~\cite{Zhan2006_K_best_sphere_decoder} for approximating the LSU estimator.  We show that this provides near sphere decoder (exact) performance, but can be computed in a reasonable amount of time if $N$ is not too large (approximately $N$ less than $1000$).  %The $K$-best method is motivated by sequential decoding algorithms and, in the context of polynomial phase estimation, bares some resemblance to the Bayesian unwrapping algorithm studied by Morelande~\cite{Morelande_bayes_unwrapping_2009_tsp}.
%We show by Montecarlo simulation that the LSU estimator approximated using the $K$-best method is statistically superior to state-of-the-art estimators from the literature including the HAF and CPF.  We also compare the computational performance of the estimators and find that BLERG.

@@

The analysis of most existing polynomial phase estimators make what is known as a `small noise' assumption~\cite{Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990}.  This amounts to assuming that the estimator $\hat{\mu}_k$ of the $k$th coefficient is such that $N^{k}(\tilde{\mu}_k -  \hat{\mu}_k) \rightarrow 0$ (in some sense) as $N \rightarrow \infty$.  This assumption is used to justify the use of first and second order expansions~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats} (typically called a `perturbation analysis' in this literature~\cite{Peleg_linear_FM,Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004}) to obtain expressions for the asymptotic mean square error of the estimator in question.  The `small noise' assumption is usually justified by Monte-Carlo simulations showing the asymptotic mean square error obtained by the perturbation analysis to be accurate provided that the signal to noise ratio (SNR) is larger than a `threshold' value (see the simulations in Section~\ref{sec:simulations} for examples).  Typically one observes the threshold to occur at smaller values of SNR as $N$ increases.  A problem with the `small noise' assumption is that it is not known that the threshold continues to occur at increasingly smaller values of SNR as $N$ increases, or if it instead converges to some finite (but perhaps small) value of SNR.  The consequence of the later occurance being that the estimator will be innaccurate if the SNR is smaller than this finite value, regardless of the number of observations $N$ obtained.  Our analysis of the LSU estimator requires no `small noise' assumpion and guarantees that no such finite value of SNR exists for the LSU estimator.  As a result the LSU estimator will be accurate for signals with arbitrarily small SNR provided that the number of observations is sufficiently large.

An analysis of the HAF estimator under the assumption that the noise $\{X_n\}$ is white and Gaussian and without the `small noise' assumption has been made in~\cite{Porat_asympt_HAF_DPT_1996} (see inparticular the proof of Lemma~1 and the remark that follows).  To date, the literature does not contain a corresponding analysis of any estimator based on phase unwrapping.  This has apparently lead to the misconception that phase unwrapping estimators are only accurate when the noise is `small' in some sense.  This misconception would appear to be motivated by early papers such as~\cite{Tretter1985,Kay1989} where it was observed that the noise occuring in the phase of the signal only looks `Gaussian' when the complex noise producing it has small variance.  The results in this paper make it clear that phase unwrapping estimators can be accurate at arbitrarily low SNR.  The information contained in the amplitude is not essential for accurate estimation of polynomial phase signals, and can be discarded with only small loss.  For example, in the case that $\{X_n\}$ is white and Gaussian, our results show that the asymptotic variance of the LSU estimator converges to the Cramer-Rao lower bound (CRB) as the SNR becomes large, and converges to approximately 3 times the CRB as the SNR approaches zero.  Related results have been presented previously~\cite{Quinn2009_dasp_phase_only_information_loss}.  It is worth noting that this factor of 3 times the CRB at low SNR is better than popular estimators such as the HAF and CPF, for which the ratio between the asymptotic variance and the CRB diverges as the SNR approaches zero~\cite{Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004,Djurovic_haf_cpf_2012}.

A key property of the LSU estimator is that it functions correctly for polynomial phase parameters contained anywhere inside the identifiable region described in Section~\ref{sec:ident_aliasing} and~\cite{Xia_dynamic_range_pps_2001,McKilliam2009IndentifiabliltyAliasingPolyphase}.  This is the maximum possible range.  By constrast popular estimators such as the HAF and CPF operate correctly only for parameters in a range much smaller than the identifiable region.  This property has been studied previously~\cite{Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Approaches exist that extend the range of parameters suitable for the HAF~\cite{Zhou_lag_diverity_PPT_1997,Djurovic_aliasing_2012}, but the resulting estimators typically have poor statistical performance.  For this reason, the LSU estimator might be the only presently existing estimator that is viable for certain applications, such as interference supression.  The theoretical results given in Section~\ref{sec:least-squar-unwr} prove that the LSU estimator functions correctly over the entire identifiable region.  To our knowledge this is the first time a result of this type has been obtained for any polynomial phase estimator.

The paper is organised in the following way. Section~\ref{sec:lattice-theory} describes some required concepts from lattice theory.  In Section~\ref{sec:ident_aliasing} we describe the identifiable region that was also derived in~\cite{McKilliam2009IndentifiabliltyAliasingPolyphase}.  These identifiability results are required in order to properly understand the statistical properties of the LSU estimator.  Section~\ref{sec:least-squar-unwr} describes the LSU estimator and states a theorem asserting the estimator to be strongly consistent and asymptotically normally distributed under some mild assumptions on the noise $\{X_n\}$. The proof of strong consistency is given in Section~\ref{sec:strongconstproof}.  Space restrictions make it impossible to give a proof of asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Section~\ref{sec:simulations} describes the results of Monte Carlo simulations that compare the performance of the LSU estimator with existing estimators including the HAF and CPF.  These simulations agree with the derived asymptotic properties. BLERG: benchmarks.

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}. The nearest lattice point problem is: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find the lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, solve
\[
\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.
\]
The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself, or rather a basis thereof, is considered as an additional input parameter \cite{micciancio_hardness_2001, AjtaiShortestVecProbNPHard1998}.  Nevertheless, algorithms exist that can compute the nearest lattice point in reasonable time if the dimension is small \cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. One such algorithm introduced by Pohst \cite{Pohst_sphere_decoder_1981} in 1981 was popularised in signal processing and communications fields by Viterbo and Boutros \cite{Viterbo_sphere_decoder_1999} and has since been called the \emph{sphere decoder}.  With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.  Approximate algorithms for computing the nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986} which has worst case polynomial complexity.  Yet another approximate algorithm is the $K$-best algorithm studied in~\cite{Zhan2006_K_best_sphere_decoder} that operates similarly to the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} used in coding theory.  The $K$-best algorithm is more expensive that Babai's algorithm, but typically produces a better approximation of the nearest lattice point.  In this paper we make use of the $K$-best algorithm for approximating the LSU estimator.  We will find that is produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount.   We will not detail the workings of this algorithm as an excellent description has been given in~\cite{Zhan2006_K_best_sphere_decoder}.

@@

It may be helpful to observe Figures~\ref{fig:circstatplot_line} and~\ref{fig:circstatplot_cube}.  In these, the phase (divided by $2\pi$) of two distinct polynomial phase signals is plotted on the left, and on the right the principal component of the phase (also divided by $2\pi$) is plotted. The circles display the samples at the integers. Note that the samples of the principal components intersect.  The corresponding polynomial phase signals are aliases.

@@

The comparitively small set of coefficients suitable for the HAF and CPF estimators has been studied previously~\cite{Angeby_PPS_aliasing_2000,Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Zhou and Wang~\cite{Zhou_lag_diverity_PPT_1997} describe a method based on the extended Euclidean algorithm that enables the HAF estimator to operate on the entire identifiable region $B$.  Unfortunately this comes at the price of statistical accuracy.  Figure~\ref{plot:m3bigzw} shows the performance of the Zhou and Wang estimator when the coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$, the same as in Figure~\ref{plot:m3big}.  The threshold for the estimator is approximately~\unit[29]{dB}.  There does not appear to exist (atleast in the literature) a similar method for increasing the range of coefficients for the CPF estimator.

@@

\begin{figure}[t] 
   	\centering 
  		\includegraphics{code/data/plot567-1.mps} 
  		\caption{BLERG} 
  		\label{plot:polyest} 
 \end{figure}

\begin{figure}[t] 
   	\centering 
  		\includegraphics{code/data/plot567-2.mps} 
  		\caption{BLERG} 
  		\label{plot:polyest} 
 \end{figure}

% \begin{figure}[p] 
%    	\centering 
%   		\includegraphics{code/data/plot567-3.mps} 
%   		\caption{BLERG} 
%   		\label{plot:polyest} 
%  \end{figure}
